The Talking Heads experiment

### 2015-04-28 chap8/chap8.tex
format relevant papers section

The Laboratorium\is{Laboratorium exhibition}
exhibition was a major artistic event in Europe during the summer of 1999 organised by 
Bruno Verbergt of an organisation called `Antwerpen Open'.
It was one of the first exhibitions that put so much emphasis on art as a research 
activity and on profound interactions between art and science.
Participants included both scientists and science historians such as Peter Galison, Bruno Latour, Israel Rosenfield
and Isabelle Stengers, as well as architects such as Rem Koolhaas and artists with an affinity for research 
such as Peter Fischli and David Weiss, Gabriel Orozco, Carsten H\"oller, Panamarenko, Lawrence Weiner and others. 
The exhibition was visited by 8000 people.

@@

Months of preparation and testing took place before we attempted to go ``live'' and public. 
Once this happened, work continued frantically to shake out bugs, get the semiotic dynamics right
and then maintain the general communication infrastructure and the ongoing interactions with participants. 
The fascinating email correspondence (see section 8.3) shows that most of the difficulties initially came 
from running the agent teleportation infrastructure. The reader has to keep in mind that this was 
late nineties, when large-scale uploading 
and downloading, cloud computing, agent architectures, etc. were totally in their infancy.

@@

A typical example of the kind of images that the cameras picked up are shown in \figref{fig:images}. 
To play a game, the agent had to find a sufficiently delineated group of objects and each object had 
to exceed a minimal size. Initially quite complex configurations were put on the white board but this made it very difficult for agents to find a coherent group and to develop good concepts for referring to the topic. The right 
image in \figref{fig:images} provides clear examples of up, down, left and right, different colors, and 
also opportunities to use multiple words (such as `red bottom').

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.40\textwidth]{chap8/figures/image-1.jpg}
\includegraphics[width=.40\textwidth]{chap8/figures/image-2.pdf}}
\caption{\label{fig:images}Examples of images picked up by the cameras. The chosen topic was signalled by drawing a bounding box around it. 
During experimentation, it was found that configurations such as in the right image lead to more stable performance.} 
\end{figure}

A typical example of an interaction is shown in \figref{fig:interaction}. 
The speaker (called ``rubber", an agent created and named by a user) has selected the blue object at the bottom left 
of the screen. We see the discrimination trees of the agent at that point and the data for each of the objects 
recognised. The coordinates of the respective objects (scaled within the coordinate system of 
the captured image) are (0.0,1.0) for the blue object, (1.0, 0.96) for the red one, and (0.42, 0.0) for the green one. 
The values are displayed as \textsc{hp} (horizontal position), \textsc{vp} (vertical position), \textsc{h} (height), \textsc{w} (width), \textsc{a} (area), 
\textsc{r} (red), \textsc{g} (green), \textsc{y} (yellow), \textsc{b} (blue), \textsc{l} (lightness). 
The blue object is chosen as the topic and the value on the blue channel has the most discriminative power, although 
there are other possibilities. There are three competing words for naming blue: `Xagadude', `Nibidesu', and `Tetipi'. 
None of them have a score higher than 0 and so a random choice of `Xagadude' is made.

The hearer (called ``me") perceives an image which is slightly different from 
the one seen by the speaker. Also the discrimination trees built up so far by this agent are different from the ones used
by the speaker. The hearer does not know the word, so the game fails. After the speaker has then pointed to the 
topic, the hearer conceptualises and guesses that the meaning of this word is `blue' because that is also for him the most 
discriminating feature of the topic (which has coordinates 0.0, 1.0) compared to the other objects in the context.

@@

An example of how the web interface displayed a single game is shown in \figref{fig:thwebsite-agent}. 
There are two agents: Antonusius is the speaker and Zelebot is the hearer. 
The green object has been chosen as topic and correctly 
recognised by the hearer using the word `kazozo'. The meaning of the word was not visible through the interface because we 
wanted users to learn themselves the language of the agents.

\begin{figure}[htb]
  \centerline{\includegraphics[width=.85\textwidth]{chap8/figures/zelebot.pdf}}
\caption{\label{fig:thwebsite-agent}Example of an interaction that took part in Antwerp. This game was a success. The agents recognised the same
set of objects and the meaning of `kazozo' was effective in finding the same topic. 
}
\end{figure}

@@

The first ``Talking Heads'' experiment ran for 4 months during the summer of 1999 and showed the validity of the mechanisms that were used for the agent architecture and of the interaction patterns and group dynamics of the agents. A shared lexicon and an underlying conceptual repertoire emerged, enabling successful communication by the agents about the scenes before them. In total, 400,000 grounded games were played. The population of agents rose to just under 2000, increasing steadily over the period of the experiment. Despite the many perturbations due to grounding, intermittent technical failures, a continuous influx of new agents entering the population, and unpredictable human interaction, the lexicon was maintained throughout this period.

@@

\figref{fig:woggle} shows the semiotic dynamics related to an expression of the meaning for the 
concept `left', i.e. a horizontal position 
between 0.0 and 0.25 (after scaling). The word ``wogglesplat" becomes dominant
although there is a very strong competition in the beginning. 
We see that users try to give other words to the same meaning, such as "gauche" or "links" (both words expressing left 
in French and German or Dutch respectively). We also see that other words such as "red" or "yellow" get associated with this \enlargethispage{1\baselineskip}
meaning, because the hearer may guess the wrong meaning in learning the word for left.

@@

\figref{fig:droite} shows another example of the semiotic dynamics in the experiment, this time looking at all the 
meanings for a particular word, namely `droite' (meaning `to the right' in French). This word clearly has been introduced
by a human user and the dominating meaning progressively becomes a region (between 0.75 and 1.0) on the horizontal position, 
as could be expected. On the way we see some confusion, Specifically there must be objects that appear both on the right 
and up (vertical position between 0.5 and 1.0).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/droite.pdf}}
\caption{\label{fig:droite}Form-Meaning diagram: Different meanings associated with the word `droite'. The dominant meaning is 
in line with the human use of the word, namely horizontal position (\textsc{hpos}) to the right of the image.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/bozopite.pdf}}
\caption{\label{fig:bozopite}Form-Meaning diagram: Different meanings associated with the word `bozopite'. 
}
\end{figure}

It was also possible that different meanings which were highly compatible were maintained in the population. 
An example is shown in the form-meaning diagram in \figref{fig:down}, which plots the average 
frequency of meanings for the word `down'. The meanings are all concepts on the vertical position channel 
\textsc{vpos}, but they carve out smaller and smaller regions.

@@

Numerous talks were given and various papers published in scientific fora. There were also various talks within 
the art context.\footnote{Examples are a `gallery talk' as Salon 3 at the Elephant and Castle Centre
in London on 2 December 1998, organised by Hans Ulrich Obrist and Molly Nesbit, and a presentation together with 
Hans-Ulrich Obrist and Rem Koolhaas in Antwerp to launch the Laboratorium book on 3 October 2001.}
The experiment received wide coverage in 
the media thanks to its public exposure as part of a major exhibition. The Süddeutsche Zeitung called it 
``Angels with Internet wings''. All this lead to invitations to show the experiment
also in other venues, as discussed in the next chapter. 
Regrettably, the challenge just to keep the experiment in the air with the available human resources
and the pressure for going on with new experiments prevented us from doing more adequate data gathering 
and analysis. Nevertheless some analyses appeared, particularly as carried out by Frederic Kaplan (\citealt{Kaplan:2001}).

@@

The experiment ended with the following ``tongue-in-cheek'' post by Angus McIntyre on the Forum:

>>>>

The Laboratorium\is{Laboratorium exhibition}
exhibition was a major artistic event in Europe during the summer of 1999 organised by 
Bruno Verbergt of an organisation called \smplenquote{Antwerpen Open}.
It was one of the first exhibitions that put so much emphasis on art as a research 
activity and on profound interactions between art and science.
Participants included both scientists and science historians such as Peter Galison, Bruno Latour, Israel Rosenfield
and Isabelle Stengers, as well as architects such as Rem Koolhaas and artists with an affinity for research 
such as Peter Fischli and David Weiss, Gabriel Orozco, Carsten H\"oller, Panamarenko, Lawrence Weiner and others. 
The exhibition was visited by 8000 people.

@@

Months of preparation and testing took place before we attempted to go \dblenquote{live} and public. 
Once this happened, work continued frantically to shake out bugs, get the semiotic dynamics right
and then maintain the general communication infrastructure and the ongoing interactions with participants. 
The fascinating email correspondence (see section 8.3) shows that most of the difficulties initially came 
from running the agent teleportation infrastructure. The reader has to keep in mind that this was 
late nineties, when large-scale uploading 
and downloading, cloud computing, agent architectures, etc. were totally in their infancy.

@@

A typical example of the kind of images that the cameras picked up are shown in \figref{fig:images}. 
To play a game, the agent had to find a sufficiently delineated group of objects and each object had 
to exceed a minimal size. Initially quite complex configurations were put on the white board but this made it very difficult for agents to find a coherent group and to develop good concepts for referring to the topic. The right 
image in \figref{fig:images} provides clear examples of up, down, left and right, different colors, and 
also opportunities to use multiple words (such as \smplenquote{red bottom}).

\begin{figure}[htbp]
 \centerline{\includegraphics[width=.40\textwidth]{chap8/figures/image-1.jpg}
\includegraphics[width=.40\textwidth]{chap8/figures/image-2.pdf}}
\caption{\label{fig:images}Examples of images picked up by the cameras. The chosen topic was signalled by drawing a bounding box around it. 
During experimentation, it was found that configurations such as in the right image led to more stable performance.} 
\end{figure}

A typical example of an interaction is shown in \figref{fig:interaction}. 
The speaker (called ``rubber", an agent created and named by a user) has selected the blue object at the bottom left 
of the screen. We see the discrimination trees of the agent at that point and the data for each of the objects 
recognised. The coordinates of the respective objects (scaled within the coordinate system of 
the captured image) are (0.0,1.0) for the blue object, (1.0, 0.96) for the red one, and (0.42, 0.0) for the green one. 
The values are displayed as \textsc{hp} (horizontal position), \textsc{vp} (vertical position), \textsc{h} (height), \textsc{w} (width), \textsc{a} (area), 
\textsc{r} (red), \textsc{g} (green), \textsc{y} (yellow), \textsc{b} (blue), \textsc{l} (lightness). 
The blue object is chosen as the topic and the value on the blue channel has the most discriminative power, although 
there are other possibilities. There are three competing words for naming blue: \smplenquote{Xagadude}, \smplenquote{Nibidesu}, and \smplenquote{Tetipi}. 
None of them have a score higher than 0 and so a random choice of \smplenquote{Xagadude} is made.

The hearer (called ``me") perceives an image which is slightly different from 
the one seen by the speaker. Also the discrimination trees built up so far by this agent are different from the ones used
by the speaker. The hearer does not know the word, so the game fails. After the speaker has then pointed to the 
topic, the hearer conceptualises and guesses that the meaning of this word is \smplenquote{blue} because that is also for him the most 
discriminating feature of the topic (which has coordinates 0.0, 1.0) compared to the other objects in the context.

@@

An example of how the web interface displayed a single game is shown in \figref{fig:thwebsite-agent}. 
There are two agents: Antonusius is the speaker and Zelebot is the hearer. 
The green object has been chosen as topic and correctly 
recognised by the hearer using the word \smplenquote{kazozo}. The meaning of the word was not visible through the interface because we 
wanted users to learn themselves the language of the agents.

\begin{figure}[htb]
  \centerline{\includegraphics[width=.85\textwidth]{chap8/figures/zelebot.pdf}}
\caption{\label{fig:thwebsite-agent}Example of an interaction that took part in Antwerp. This game was a success. The agents recognised the same
set of objects and the meaning of \smplenquote{kazozo} was effective in finding the same topic. 
}
\end{figure}

@@

The first \dblenquote{Talking Heads} experiment ran for 4 months during the summer of 1999 and showed the validity of the mechanisms that were used for the agent architecture and of the interaction patterns and group dynamics of the agents. A shared lexicon and an underlying conceptual repertoire emerged, enabling successful communication by the agents about the scenes before them. In total, 400,000 grounded games were played. The population of agents rose to just under 2000, increasing steadily over the period of the experiment. Despite the many perturbations due to grounding, intermittent technical failures, a continuous influx of new agents entering the population, and unpredictable human interaction, the lexicon was maintained throughout this period.

@@

\figref{fig:woggle} shows the semiotic dynamics related to an expression of the meaning for the 
concept \smplenquote{left}, i.e. a horizontal position 
between 0.0 and 0.25 (after scaling). The word ``wogglesplat" becomes dominant
although there is a very strong competition in the beginning. 
We see that users try to give other words to the same meaning, such as "gauche" or "links" (both words expressing left 
in French and German or Dutch respectively). We also see that other words such as "red" or "yellow" get associated with this \enlargethispage{1\baselineskip}
meaning, because the hearer may guess the wrong meaning in learning the word for left.

@@

\figref{fig:droite} shows another example of the semiotic dynamics in the experiment, this time looking at all the 
meanings for a particular word, namely \smplenquote{droite} (meaning \smplenquote{to the right} in French). This word clearly has been introduced
by a human user and the dominating meaning progressively becomes a region (between 0.75 and 1.0) on the horizontal position, 
as could be expected. On the way we see some confusion, Specifically there must be objects that appear both on the right 
and up (vertical position between 0.5 and 1.0).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/droite.pdf}}
\caption{\label{fig:droite}Form-Meaning diagram: Different meanings associated with the word \smplenquote{droite}. The dominant meaning is 
in line with the human use of the word, namely horizontal position (\textsc{hpos}) to the right of the image.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap8/figures/bozopite.pdf}}
\caption{\label{fig:bozopite}Form-Meaning diagram: Different meanings associated with the word \smplenquote{bozopite}. 
}
\end{figure}

It was also possible that different meanings which were highly compatible were maintained in the population. 
An example is shown in the form-meaning diagram in \figref{fig:down}, which plots the average 
frequency of meanings for the word \smplenquote{down}. The meanings are all concepts on the vertical position channel 
\textsc{vpos}, but they carve out smaller and smaller regions.

@@

Numerous talks were given and various papers published in scientific fora. There were also various talks within 
the art context.\footnote{Examples are a \smplenquote{gallery talk} as Salon 3 at the Elephant and Castle Centre
in London on 2 December 1998, organised by Hans Ulrich Obrist and Molly Nesbit, and a presentation together with 
Hans-Ulrich Obrist and Rem Koolhaas in Antwerp to launch the Laboratorium book on 3 October 2001.}
The experiment received wide coverage in 
the media thanks to its public exposure as part of a major exhibition. The Süddeutsche Zeitung called it 
\dblenquote{Angels with Internet wings}. All this led to invitations to show the experiment
also in other venues, as discussed in the next chapter. 
Regrettably, the challenge just to keep the experiment in the air with the available human resources
and the pressure for going on with new experiments prevented us from doing more adequate data gathering 
and analysis. Nevertheless some analyses appeared, particularly as carried out by Frederic Kaplan (\citealt{Kaplan:2001}).

@@

The experiment ended with the following \dblenquote{tongue-in-cheek} post by Angus McIntyre on the Forum:

### 2015-04-28 chap9/chap9.tex
format relevant papers section

``The Kettle's Yard Gallery is associated with Cambridge University. It acted as the main site of the N01SE exhibition 
and showed a variety of historical artefacts (including the brain of Babbage and the original DNA structure built by 
Watson and Crick) together with new artistic works. The catalogueue published by the Kettle's Yard 
Gallery \footnote{www.kettlesyard.co.uk/noise} featured articles by Brian Smith, Umberto Eco, Bruno Latour, 
Bruce Sterling, Luc Steels, Peter Weibel and others. Below is my text as it appeared in the catalogue: 
\begin{quotation}
Meanings are not a priori Platonic entities independent of language; meanings are the result of embodied interactions with the world, obtained via the role words play in verbal interactions called ``language games''. The Talking Heads Experiment explores one kind of language game: a guessing game played by two robotic agents about the scene directly in front of them. One agent acts as Speaker and attempts to draw the attention of the Hearer to some object by transmitting a verbal description of it; the Speaker succeeds if the Hearer correctly guesses which object is ``meant''.

@@

The installation itself was similar to the one in Cambridge. There was a wall with 
geometric figures pasted on it, posters explaining the exhibition, the two pan-tilt cameras mounted on tripods, 
and the computers driving the software (see \figref{fig:london-heads}). The London site posed particularly hard 
problems in the alignment of the cameras. 
It turned out that a subway was passing under the gallery and causing strong vibrations 
every few minutes, which caused the cameras to physically shift on the floor. Because the pointing behaviour was sensitive 
to alignment and prior calibration, this lead to growing pointing errors and 
subsequent errors in feedback, causing a strong decline in 
the success rate and occasional chaos in the agents' vocabularies. This was partially offset by stable conditions in 
other sites but nevertheless made the task of reaching coherence virtually impossible.

@@

To deal with the problem of agent congestion, software issues, and camera misalignment at the London 
and Cambridge sites, maintenance was carried out which lead to a temporary unavailability of the experiment. 
Surprisingly, the reaction of the Hull group was entirely negative as they saw it as a 
way to prevent them from gaining or keeping control of the language used by the agents, even though 
Angus McIntyre clearly and patiently explained why maintenance was necessary. Also, Denna Jones ({\itshape Phlox}) took 
it personally as she did not realise that the calibration errors were due to vibrations caused by
traffic of the London Underground near and under the Wellcome Gallery.

@@

The {\itshape Mus\'ee d'Art Moderne} in Paris organised a solo exhibition of artist Olafur Eliasson entitled 
``Chaque matin je me sens diff\'erent. Chaque soir je me sens le m\^eme" between 
22 March and 12 May 2002. (\citealt{Scherf:2002}) Olafur Eliasson is known for his thorough 
investigations of colour, such as using monochromatic light to create an artificial sun in the Modern Tate
Gallery in London (2003). Eliasson invited me to jointly work out a new interpretation of the 
Talking Heads Experiment, which became ``Look into the Box''. Nicolas Neubauer, a master student working at 
that time at the Sony Computer Science Laboratory, was the chief implementer together with Angus McIntyre.  
The set-up and results are described in detail in \cite{Steels:2004} and \cite{Neubauer:2004}.\is{Museum of Modern Art installation}

>>>>

``The Kettle's Yard Gallery is associated with Cambridge University. It acted as the main site of the N01SE exhibition 
and showed a variety of historical artefacts (including the brain of Babbage and the original DNA structure built by 
Watson and Crick) together with new artistic works. The catalogueue published by the Kettle's Yard 
Gallery \footnote{www.kettlesyard.co.uk/noise} featured articles by Brian Smith, Umberto Eco, Bruno Latour, 
Bruce Sterling, Luc Steels, Peter Weibel and others. Below is my text as it appeared in the catalogue: 
\begin{quotation}
Meanings are not a priori Platonic entities independent of language; meanings are the result of embodied interactions with the world, obtained via the role words play in verbal interactions called ``language games''. The Talking Heads experiment explores one kind of language game: a guessing game played by two robotic agents about the scene directly in front of them. One agent acts as Speaker and attempts to draw the attention of the Hearer to some object by transmitting a verbal description of it; the Speaker succeeds if the Hearer correctly guesses which object is ``meant''.

@@

The installation itself was similar to the one in Cambridge. There was a wall with 
geometric figures pasted on it, posters explaining the exhibition, the two pan-tilt cameras mounted on tripods, 
and the computers driving the software (see \figref{fig:london-heads}). The London site posed particularly hard 
problems in the alignment of the cameras. 
It turned out that a subway was passing under the gallery and causing strong vibrations 
every few minutes, which caused the cameras to physically shift on the floor. Because the pointing behaviour was sensitive 
to alignment and prior calibration, this led to growing pointing errors and 
subsequent errors in feedback, causing a strong decline in 
the success rate and occasional chaos in the agents' vocabularies. This was partially offset by stable conditions in 
other sites but nevertheless made the task of reaching coherence virtually impossible.

@@

To deal with the problem of agent congestion, software issues, and camera misalignment at the London 
and Cambridge sites, maintenance was carried out which led to a temporary unavailability of the experiment. 
Surprisingly, the reaction of the Hull group was entirely negative as they saw it as a 
way to prevent them from gaining or keeping control of the language used by the agents, even though 
Angus McIntyre clearly and patiently explained why maintenance was necessary. Also, Denna Jones ({\itshape Phlox}) took 
it personally as she did not realise that the calibration errors were due to vibrations caused by
traffic of the London Underground near and under the Wellcome Gallery.

@@

The {\itshape Mus\'ee d'Art Moderne} in Paris organised a solo exhibition of artist Olafur Eliasson entitled 
``Chaque matin je me sens diff\'erent. Chaque soir je me sens le m\^eme" between 
22 March and 12 May 2002. (\citealt{Scherf:2002}) Olafur Eliasson is known for his thorough 
investigations of colour, such as using monochromatic light to create an artificial sun in the Modern Tate
Gallery in London (2003). Eliasson invited me to jointly work out a new interpretation of the 
Talking Heads experiment, which became ``Look into the Box''. Nicolas Neubauer, a master student working at 
that time at the Sony Computer Science Laboratory, was the chief implementer together with Angus McIntyre.  
The set-up and results are described in detail in \cite{Steels:2004} and \cite{Neubauer:2004}.\is{Museum of Modern Art installation}

### 2015-04-28 ind-steelsbook.tex
fixed remaining widows and orphans
first pass of bibliography
format relevant papers section

\proofreader{Martin Hilpert, Felix Kopecky, Peter Petré, Daniela Schröder}
\typesetter{Felix Kopecky, Sebastian Nordhoff}
\title{The Talking Heads Experiment}
\subtitle{Origins of words and meanings}
\BackTitle{The Talking Heads Experiment}% - \\ Origins of Words and Meanings}
\author{Luc Steels}
%\Affiliation{ Catalan Institute for Advanced Studies (ICREA) \\ %- Instituci\accute{o} Catalana de Recerca i Estudis Avan\cats\\
%Institut de Biologia Evolutiva (Universitat Pompeu Fabra and CSIC) Barcelona}
\dedication{For my daughter Lenie}
\BackBody{The Talking Heads Experiment, conducted in the years 1999--2001, was the first large-scale experiment 
in which open populations of situated embodied agents created for the first time ever a new shared vocabulary
by playing language games about real world scenes in front of them. The agents could teleport to different 
physical sites in the world through the Internet. Sites, in Antwerp, Brussels, Paris, Tokyo, London, Cambridge
and several other locations were linked into the network. Humans could interact with the robotic agents either
on site or remotely through the Internet and thus influence the evolving ontologies and languages of the 
artificial agents.

@@

%\thispagestyle{empty}
\begin{figure}[htbp]
  \centerline{\includegraphics[width=.95\textwidth]{extra/figs/cover-original-book.pdf}}
\caption{\small\label{left-right} Original cover of the Talking Heads book published in 1999 on the occasion of the 
Laboratorium exhibition in Antwerp.}
\label{f:cover}
\end{figure}
%
\include{chap1/chap1}
\include{chap2/chap2}
\include{chap3/chap3}
\include{chap4/chap4}
\include{chap5/chap5}
\include{chap6/chap6}
\include{chap7/chap7}
%
\part{Installations and experiments} \label{part:2}
\include{chap8/chap8}
\include{chap9/chap9}

@@

% \part{Appendices} \label{part:4}
\include{extra/papers-language-steels}

% \phantomsection%this allows hyperlink in ToC to work
% \addcontentsline{toc}{chapter}{List of references}
\printbibliography %change to the name of your bib file
\clearpage

\addcontentsline{toc}{chapter}{Index} 
\phantomsection%this allows hyperlink in ToC to work
\addcontentsline{toc}{section}{Name index}
\ohead{Name index} 
\printindex

>>>>

\proofreader{Martin Hilpert, Felix Kopecky, Peter Petré, Daniela Schröder}
\typesetter{Felix Kopecky, Sebastian Nordhoff}
\title{The Talking Heads experiment}
\subtitle{Origins of words and meanings}
\BackTitle{The Talking Heads experiment}% - \\ Origins of Words and Meanings}
\author{Luc Steels}
%\Affiliation{ Catalan Institute for Advanced Studies (ICREA) \\ %- Instituci\accute{o} Catalana de Recerca i Estudis Avan\cats\\
%Institut de Biologia Evolutiva (Universitat Pompeu Fabra and CSIC) Barcelona}
\dedication{For my daughter Lenie}
\BackBody{The Talking Heads experiment, conducted in the years 1999--2001, was the first large-scale experiment 
in which open populations of situated embodied agents created for the first time ever a new shared vocabulary
by playing language games about real world scenes in front of them. The agents could teleport to different 
physical sites in the world through the Internet. Sites, in Antwerp, Brussels, Paris, Tokyo, London, Cambridge
and several other locations were linked into the network. Humans could interact with the robotic agents either
on site or remotely through the Internet and thus influence the evolving ontologies and languages of the 
artificial agents.

@@

%\thispagestyle{empty}
\begin{figure}[htbp]
  \centerline{\includegraphics[width=.95\textwidth]{extra/figs/cover-original-book.pdf}}

{ Original cover of the Talking Heads book published in 1999 on the occasion of the 
Laboratorium exhibition in Antwerp.}
% \label{f:cover}
\end{figure}
% %
\include{chap1/chap1}
\include{chap2/chap2}
\include{chap3/chap3}
\include{chap4/chap4}
\include{chap5/chap5}
\include{chap6/chap6}
\include{chap7/chap7}
%
\part{Installations and experiments} \label{part:2}
\include{chap8/chap8}
\include{chap9/chap9}

@@

% \part{Appendices} \label{part:4}
% \include{extra/papers-language-steels}
% \input{extra/steelspapers.bbl}

\phantomsection%this allows hyperlink in ToC to work
\addchap{Relevant papers by Luc Steels}
\phantomsection

\renewcommand{\bibitem}[2][]{\item}
\newcommand{\natexlab}[1]{}
\newcommand{\doi}[1]{#1}
\newcommand{\urlprefix}[1]{#1}
\begin{enumerate}
\input{extra/steelspapers.bbl}
\end{enumerate}

\phantomsection
\printbibliography %change to the name of your bib file
\clearpage

\phantomsection%this allows hyperlink in ToC to work
\addcontentsline{toc}{chapter}{Index} 
\addcontentsline{toc}{section}{Name index}
\ohead{Name index} 
\printindex

### 2015-04-28 preface/preface.tex
format relevant papers section

I set up the Talking Heads Experiment with a group of brilliant students and collaborators at the end of the nineteen nineties. It was intended to be the first large-scale, open-ended experiment in the emergence of a shared set of grounded concepts and a vocabulary for expressing these concepts by a population of autonomous agents. Inspired by Ludwig Wittgenstein, the experiment took the form of a series of language games, more concretely games of reference about a ``world'' made up of geometric figures pasted on a white board and observable by the agents through pan-tilt cameras. 
I  wanted to demonstrate with this experiment earlier breakthroughs in the study of language origins and 
test whether they would hold for large-scale populations and open-ended environments. I also wanted to find 
out how humans would interact with these agents. So we made it such that human users, after logging in through the 
Internet, could teach new words to agents or use the words they learned from the agents to play their own language games.

@@

At the occasion of the 1999 Laboratorium exhibition, the draft of a book was published 
that described the experiment and the underlying theoretical assumptions in considerable detail. 
For many reasons, not at least 
that work continued at great speed on other exciting experiments, the original ``pre-edition'' of the
book never made it to a fully finished officially 
published work, and circulated only as an ``underground'' edition. 
This was disappointing because the Talking Heads Experiment was an important breakthrough. 
Moreover the experiment contained the first inklings of mechanisms that since 
then have been worked out, enhanced and tested in many experiments which replicated the original results and further
enhanced them. The present volume is intended to fill this gap.

@@

The present new edition is not only special because of the historical significance of the Talking Heads 
Experiment but also because it is the first one in a new series ``Computational Models of 
Language Evolution'' published by the Language Science Press. The intention of this series is to make available 
through Open Access in-depth models of language evolution that have been validated using agent-based 
computational simulations. I am grateful to Martin Haspelmath and Stefan M\"uller (the editors of the Language Science Press)
for making it possible to publish in Open Access research results which do not quite fit in the standard mode. I thank 
Maria Ferrer Bonnet for help in the final stage of adapting bibliographies. And I am grateful to ICREA for time to 
create a revised and extended version of the Talking Heads Book and to the Institut de Biologia Evolutiva in Barcelona 
for providing such an excellent working environment. I thank in addition Remi van Trijp for his help in making this series a reality 
and Annemie Maes for her encouragement in the final stretches to finish this book.

>>>>

I set up the Talking Heads experiment with a group of brilliant students and collaborators at the end of the nineteen nineties. It was intended to be the first large-scale, open-ended experiment in the emergence of a shared set of grounded concepts and a vocabulary for expressing these concepts by a population of autonomous agents. Inspired by Ludwig Wittgenstein, the experiment took the form of a series of language games, more concretely games of reference about a ``world'' made up of geometric figures pasted on a white board and observable by the agents through pan-tilt cameras. 
I  wanted to demonstrate with this experiment earlier breakthroughs in the study of language origins and 
test whether they would hold for large-scale populations and open-ended environments. I also wanted to find 
out how humans would interact with these agents. So we made it such that human users, after logging in through the 
Internet, could teach new words to agents or use the words they learned from the agents to play their own language games.

@@

At the occasion of the 1999 Laboratorium exhibition, the draft of a book was published 
that described the experiment and the underlying theoretical assumptions in considerable detail. 
For many reasons, not at least 
that work continued at great speed on other exciting experiments, the original ``pre-edition'' of the
book never made it to a fully finished officially 
published work, and circulated only as an ``underground'' edition. 
This was disappointing because the Talking Heads experiment was an important breakthrough. 
Moreover the experiment contained the first inklings of mechanisms that since 
then have been worked out, enhanced and tested in many experiments which replicated the original results and further
enhanced them. The present volume is intended to fill this gap.

@@

The present new edition is not only special because of the historical significance of the Talking Heads 
experiment but also because it is the first one in a new series ``Computational Models of 
Language Evolution'' published by the Language Science Press. The intention of this series is to make available 
through Open Access in-depth models of language evolution that have been validated using agent-based 
computational simulations. I am grateful to Martin Haspelmath and Stefan M\"uller (the editors of the Language Science Press)
for making it possible to publish in Open Access research results which do not quite fit in the standard mode. I thank 
Maria Ferrer Bonnet for help in the final stage of adapting bibliographies. And I am grateful to ICREA for time to 
create a revised and extended version of the Talking Heads Book and to the Institut de Biologia Evolutiva in Barcelona 
for providing such an excellent working environment. I thank in addition Remi van Trijp for his help in making this series a reality 
and Annemie Maes for her encouragement in the final stretches to finish this book.
